Continuous gravitational waves in the lab: recovering audio signals with a table-top optical microphone

### 2020-08-16 paper/ifo-complexAudio.tex
some general text editing and addressing comments

\begin{figure*}
	\includegraphics[width=.6\textwidth]{figures/setup_pic2.pdf}
\caption{Photograph of the optical microphone. The Michelson interferometer is shown on the left and the circuit and Raspberry Pi on the right. The interferometer is laid out as in Figure~\ref{fig:ifo_schematic_podo}. In the main image, the photodiode is placed behind the cloth screen. The inset shows a face-on view of the photodiode with the covering screen removed. See Appendix~\ref{app:circuit_diagram} for a circuit diagram and detailed photograph of the circuit.}
	\label{fig:setup_pic2}
\end{figure*}

@@

\begin{figure*}
\begin{center}
% trim order is <left> <lower> <right> <upper>
\includegraphics[width=\textwidth,trim={0cm 8.5cm 0 8cm},clip]{figures/freqSpectrumOriginalNoiseSubtracted.pdf}
\end{center}
\caption{\label{fig:noiseSubtract}
Frequency spectrum of a $1\,{\rm s}$ signal before the use of filters. 
The first panel shows the original frequency spectrum of the human voice recording (saying ``A cathode ...''). 
The second panel shows the spectrum recorded after the signal passes through the optical microphone (sampled at $16\,000\,{\rm Hz}$).
The third panel shows a noise-only recording from the optical microphone when no audio signal is being played. 
The fourth panel shows the result of subtracting the noise-only spectrum in the third panel from the recording in the second panel. 
We see no obvious improvement from this method. 
}
\end{figure*}

@@

\begin{figure*}
\begin{center}
\includegraphics[width=\textwidth,trim={0cm 8.7cm 0 8.3cm},clip]{figures/timeSeriesNotchWiener.pdf}
\end{center}
\caption{\label{fig:timeNotchWiener}
Timeseries results for the notch and Wiener filters. 
The first panel shows the original speech recording which was played from the speaker. 
The second panel shows the data from the optical microphone before any filters are applied. 
The third and fourth panels show results after applying the notch and Wiener filter respectively. 
Finally, the fifth panel shows the result of applying both the notch and Wiener filter. 
}
\end{figure*}

\begin{figure*}
\begin{center}
\includegraphics[width=\textwidth,trim={0cm 8.2cm 0 8cm},clip]{figures/freqNotchWiener.pdf}
\end{center}
\caption{\label{fig:freqNotchWiener}
Frequency spectrum results for the notch and Wiener filters. 
The first and second panels show the results of applying the notch and Wiener filter respectively.
The third panel shows the spectrum after applying both the notch and Wiener filters. 
The first, second, and third panels show the spectrum of the timeseries show in the third, fourth, and fifth panels of Fig.~\ref{fig:timeNotchWiener} respectively. 
}
\end{figure*}

>>>>

\begin{figure*}
	\includegraphics[width=.7\textwidth]{figures/setup_pic2.pdf}
\caption{Photograph of the optical microphone. The Michelson interferometer is shown on the left and the circuit and Raspberry Pi on the right. The interferometer is laid out as in Figure~\ref{fig:ifo_schematic_podo}. In the main image, the photodiode is placed behind the cloth screen. The inset shows a face-on view of the photodiode with the covering screen removed. See Appendix~\ref{app:circuit_diagram} for a circuit diagram and detailed photograph of the circuit.}
	\label{fig:setup_pic2}
\end{figure*}

@@

\begin{figure*}
\begin{center}
% trim order is <left> <lower> <right> <upper>
\includegraphics[width=\textwidth,trim={1cm 8.5cm 1cm 8cm},clip]{figures/freqSpectrumOriginalNoiseSubtracted.pdf}
\end{center}
\caption{\label{fig:noiseSubtract}
Frequency spectrum of a $1\,{\rm s}$ signal before the use of filters. 
The first panel shows the original frequency spectrum of the human voice recording (saying ``A cathode ...''). 
The second panel shows the spectrum recorded after the signal passes through the optical microphone (sampled at $16\,000\,{\rm Hz}$).
The third panel shows a noise-only recording from the optical microphone when no audio signal is being played. 
The fourth panel shows the result of subtracting the noise-only spectrum in the third panel from the recording in the second panel. 
We see no obvious improvement from this method. 
}
\end{figure*}

@@

\begin{figure*}
\begin{center}
\includegraphics[width=\textwidth,trim={1cm 8.2cm 1 8.3cm},clip]{figures/timeSeriesNotchWiener.pdf}
\end{center}
\caption{\label{fig:timeNotchWiener}
Timeseries results for the notch and Wiener filters. 
The first panel shows the original speech recording which was played from the speaker. 
The second panel shows the data from the optical microphone before any filters are applied. 
The third and fourth panels show results after applying the notch and Wiener filter respectively. 
Finally, the fifth panel shows the result of applying both the notch and Wiener filter. 
}
\end{figure*}

\begin{figure*}
\begin{center}
\includegraphics[width=\textwidth,trim={0cm 8.2cm 0cm 8cm},clip]{figures/freqNotchWiener.pdf}
\end{center}
\caption{\label{fig:freqNotchWiener}
Frequency spectrum results for the notch and Wiener filters. 
The first and second panels show the results of applying the notch and Wiener filter respectively.
The third panel shows the spectrum after applying both the notch and Wiener filters. 
The first, second, and third panels show the spectrum of the timeseries show in the third, fourth, and fifth panels of Fig.~\ref{fig:timeNotchWiener} respectively. 
}
\end{figure*}

### 2020-08-16 paper/ifo-introduction.tex
some general text editing and addressing comments

% general gw intro
In 2015, gravitational waves were observed for the first time from the merger of two black holes~\cite{GW150914}. 
The observation, made by the Laser Interferometer Gravitational-wave Observatory~\citep[LIGO]{AdvancedLIGO:2015}, marked a breakthrough in modern astrophysics and revealed a new means to observe the universe. 
Since 2015, the LIGO and Virgo~\cite{AdvancedVirgo:2015} observatories have made numerous detections of binary black hole~\cite{GW151226,GW170104,GW170814} and binary neutron star~\cite{GW170817,GW170817multi,GW190425} mergers. 
Gravitational-wave detection has led to increased public and media interest in gravitational-wave science. 
Gravitational-wave research groups around the world have produced demonstrations and activities to explain their research to general audiences. 
Activities range from hands-on demonstrations, exhibitions~\cite{TTExhibit:2020,LIGOScienceEducationCentre:online,GravityDiscoveryCentre:online}, and online data analysis tutorials~\cite{GWOSC:online,LOSC:2015}, to phone apps~\cite{LaserLabs:online,SciVR:online,chirp:online}, online games~\cite{BlackHoleHunter:online}, and musical performances~\cite{ArthurJeffesMusic:online,GravitySynthLeonTrimble:online}%\han{look up Sarah Farmer's project}.

% what are gws and how do we detect them...
Gravitational waves are a prediction of Albert Einstein's theory of general relativity~\cite{Einstein:1916}. 
They are disturbances in spacetime caused by the acceleration of asymmetric massive objects. 
The effect of gravitational waves is a change in lengths: a `stretching and squeezing' of the distance between objects.
Observatories such as LIGO, Virgo, and KAGRA~\cite{KAGRA:2013} are laser interferometers.
They use the interference of laser light to measure changes in distance. 
%As shown in an aerial view of LIGO Hanford in Figure~\ref{fig:ligo_pic}, they use beam arms kilometres long.
Gravitational-wave observatories are extremely complex but are fundamentally based on the Michelson interferometer. 
Table-top Michelson interferometers are commonly used in undergraduate lab experiments~\cite{UgoliniEtAl:2019} and to demonstrate the science of gravitational wave detection to non-specialist audiences~\cite{ThorLabsIFO,NikhefIFO,TTExhibit:2020,LIGOIFOGlue,LIGOIFOMagnets}.

% continuous gws
To date the network of gravitational wave observatories has observed short-duration transient signals~\cite{GWTC-1:2018,GWOSC:online}. 
However, the network is also searching for continuous gravitational waves: persistent, periodic, near-monochromatic signals.
Continuous waves may be emitted at constant frequencies, or may wander slowly in frequency over time. 
For example, rotating neutron stars in low mass X-ray binaries (LMXBs) are one of the targets for continuous-wave searches. 
LMXBs are binary systems containing a compact object (such as a neutron star or a black hole) in orbit with a low mass stellar companion, where the rotation frequency of the compact object wanders over time due to variable accretion of matter (and hence angular momentum transfer) from the companion~\cite{xraybinaries:1997}. 
Scorpius X-1 is an LMXB which is particularly bright in X-rays. 
It is a prime target for continuous wave searches from LMXBs and numerous, as yet unsuccessful, searches have been performed~\cite{ScoX1O2Viterbi:2019,RadiometerO1O2:2019,SearchCrossCorrO1:2017}.

@@

The paper presents 
In this work we present a selection of demonstrations using different types of audio signals ranging from simple tones, analogous to continuous wave signals, to more complex audio of music and speech. 
In Section~\ref{sec:ifo} we detail the table-top interferometer design. 
We demonstrate observing a single note from a speaker in Section~\ref{sec:single_tone}.
In Section~\ref{sec:viterbi_wandering} we replace the single note with a slowly changing frequency and analyse the signal using a hidden Markov model technique from continuous-wave searches. 
In Section~\ref{sec:optical_microphone}, we demonstrate capturing and playing back complex audio, such as music and speech. 
This demonstration of an `optical microphone' serves as a more general exhibition of signal processing. 
We suggest avenues of future work in Section~\ref{sec:future_work} and draw conclusions in Section~\ref{sec:conclusions}.
Appendix~\ref{app:code} collects the software and scripts used to produce this work.

>>>>

% general gw intro
In 2015, gravitational waves were observed for the first time from the merger of two black holes~\cite{GW150914}. 
The observation, made by the Advanced Laser Interferometer Gravitational-wave Observatory~(Advanced LIGO \citep{AdvancedLIGO:2015}), marked a breakthrough in modern astrophysics and revealed a new means to observe the universe. 
Since 2015, the Advanced LIGO and Advanced Virgo~\cite{AdvancedVirgo:2015} observatories have made numerous detections of binary black hole~\cite{GW151226,GW170104,GW170814} and binary neutron star~\cite{GW170817,GW170817multi,GW190425} mergers. 
Gravitational-wave detection has led to increased public and media interest in gravitational-wave science. 
Gravitational-wave research groups around the world have produced demonstrations and activities to explain their research to general audiences. 
Activities range from hands-on demonstrations, exhibitions~\cite{TTExhibit:2020,LIGOScienceEducationCentre:online,GravityDiscoveryCentre:online}, and online data analysis tutorials~\cite{GWOSC:online,LOSC:2015}, to phone apps~\cite{LaserLabs:online,SciVR:online,chirp:online}, online games~\cite{BlackHoleHunter:online}, and musical performances~\cite{ArthurJeffesMusic:online,GravitySynthLeonTrimble:online}%\han{look up Sarah Farmer's project}.

% what are gws and how do we detect them...
Gravitational waves are a prediction of Albert Einstein's theory of general relativity~\cite{Einstein:1916}. 
They are disturbances in spacetime caused by the acceleration of asymmetric massive objects. 
The effect of gravitational waves is a change in lengths: a `stretching and squeezing' of the distance between objects.
Observatories such as LIGO, Virgo, GEO600~\cite{GEO600:2010}, and KAGRA~\cite{KAGRA:2013} are laser interferometers.
They use the interference of laser light to measure changes in distance. 
%As shown in an aerial view of LIGO Hanford in Figure~\ref{fig:ligo_pic}, they use beam arms kilometres long.
Gravitational-wave observatories are extremely complex but are fundamentally based on the Michelson interferometer. 
Table-top Michelson interferometers are commonly used in undergraduate lab experiments~\cite{UgoliniEtAl:2019} and to demonstrate the science of gravitational wave detection to non-specialist audiences~\cite{ThorLabsIFO,NikhefIFO,AMIGO:online,TTExhibit:2020,LIGOIFOGlue,LIGOIFOMagnets}.

% continuous gws
To date the network of gravitational-wave observatories has observed short-duration transient signals~\cite{GWTC-1:2018,GWOSC:online}. 
However, the network is also searching for continuous gravitational waves: persistent, periodic, near-monochromatic signals.
Continuous waves may be emitted at constant frequencies, or may wander slowly in frequency over time. 
For example, rotating neutron stars in low mass X-ray binaries (LMXBs) are one of the targets for continuous-wave searches. 
LMXBs are binary systems containing a compact object (such as a neutron star or a black hole) in orbit with a low mass stellar companion, where the rotation frequency of the compact object wanders over time due to variable accretion of matter (and hence angular momentum transfer) from the companion~\cite{xraybinaries:1997}. 
Scorpius X-1 is an LMXB which is particularly bright in X-rays making it a prime target for continuous wave searches from LMXBs. 
Numerous, as yet unsuccessful, searches have been performed for Scorpius X-1 and other LMXBs~\cite{ScoX1O2Viterbi:2019,RadiometerO1O2:2019,SearchCrossCorrO1:2017,ScoX1ViterbiO1:2017,SearchRadiometerO1:2017,SearchCrossCorrO1:2017,MiddletonEtAlO2LMXBs:2020,SearchTwoSpecS6:2017}.

@@

In this work we present a selection of demonstrations using different types of audio signals ranging from simple tones, analogous to continuous wave signals, to more complex audio of music and speech. 
In Section~\ref{sec:ifo} we detail the table-top interferometer design. 
We demonstrate observing a single note from a speaker in Section~\ref{sec:single_tone}.
In Section~\ref{sec:viterbi_wandering} we replace the single note with a slowly changing frequency and analyse the signal using a hidden Markov model technique from continuous-wave searches. 
In Section~\ref{sec:optical_microphone}, we demonstrate capturing and playing back complex audio, such as music and speech. 
This demonstration of an `optical microphone' serves as a more general exhibition of signal processing. 
We suggest avenues of future work in Section~\ref{sec:future_work} and draw conclusions in Section~\ref{sec:conclusions}.
Appendix~\ref{app:code} collects the software and scripts used to produce this work.

### 2020-08-16 paper/ifo-tableTopGWs.tex
some general text editing and addressing comments

Gravitational-wave detectors are large, complex interferometers.
The LIGO and Virgo detectors have $4\,{\rm km}$ and $3\,{\rm km}$ arms respectively.
They are sensitive to minuscule changes in distance, for example the first detection of a binary black hole merger had a peak gravitational-wave strain of $10^{-21}$ ~\cite{GW150914}.
However, their design is based fundamentally on the Michelson interferometer. 
Detection and analysis concepts can be demonstrated with table-top optical equipment. 
See Refs.~\cite{TTExhibit:2020,TTExhibit:online,ThorLabsIFO,LIGOIFOGlue,LIGOIFOMagnets,FoxEtAl:1999} for a selection of table-top Michelson interferometer designs used for science communication.

>>>>

Gravitational-wave detectors are large, complex interferometers.
The LIGO and Virgo detectors have $4\,{\rm km}$ and $3\,{\rm km}$ arms respectively.
They are sensitive to minuscule changes in distance, for example the first detection of a binary black hole merger had a peak gravitational-wave strain of $10^{-21}$ ~\cite{GW150914}.
However, their design is based fundamentally on the Michelson interferometer. 
Detection and analysis concepts can be demonstrated with table-top optical equipment. 
See Refs.~\cite{TTExhibit:2020,TTExhibit:online,AMIGO:online,ThorLabsIFO,LIGOIFOGlue,LIGOIFOMagnets,FoxEtAl:1999} for a selection of table-top Michelson interferometer designs used for science communication.

### 2020-08-16 paper/ifo-wanderingTone.tex
task 2 => tasks complete
some general text editing and addressing comments

% Tracking tones with the Viterbi algorithm
% explaination and successful application
% lean on the continuous gravitational waves angle
Continuous gravitational wave signals may wander slowly in frequency over time, due to stochastic internal processes in the superfluid interior of isolated neutron stars~\cite{MelatosDouglassSimula:2015,Jones:2010} or stochastic accretion flows in binary neutron stars~\cite{BildstenTB:1998}. 
%As also mentioned in Section~\ref{sec:introduction}, continuous wave searches often look for signals that wander slowly in frequency~\cite{ScoX1O2Viterbi:2019}. 
The audio analogue of this is a tone that wanders in frequency; a note that changes pitch. 
%Unlike the single tone described in Section~\ref{sec:single_tone}, finding the signal isn’t as simple as searching for a peak in the spectrum for the full timeseries data.%observing run.

@@

%The data from a gravitaitonal wave observation run can be many months to a year. 
%One method to search for a slowly wandering continuous wave signal is to split the data into several shorter intervals. 
%Each chunk is analysed seperately. 
%In our project we take a Fourier transform of each interval and then create a grid (or spectrogram) where each cell is the Fourier amplitude of a particular frequency at a particular time. 
%(In continuous wave searches, a detection statistic is used to calculate the likelihood of a signal being present in a particular frequency bin.)
%The task is then to find the best path through the spectrogram grid as if it where a weighted graph, where here the weights are the Fourier amplitudes. 
%The Viterbi algorithm~\cite{Viterbi:1967} is an efficient method find the best path through the grid. 
%It is used in contiuous wave searches for a range of astrophysical targets~\cite{ScoX1O2Viterbi:2019,ScoX1ViterbiO1:2017,MillhouseStrangMelatos:2020,PostMergerRemnantSearch:2019,SunEtAlSNR:2018,viterbi_application}.

@@

\item The Viterbi path is then found by backtracking along the best paths at each time step.
\jam{[Andrew mentioned providing a closed form here, which I'm not sure exists. The code works by storing three arrays: the weights, the score for the best path to a given vertex, and the index of the previous vertex along that path. Backtracking comprises of looping through that third array from the vertex in the last row with the highest score from the second array. I don't think we need to explain it any further than that.]}
In Fig.~\ref{fig:viterbi}, we see that the path ending at (b) started at (c). 
The Viterbi path is highlighted in orange in fig.~\ref{fig:viterbi}.
\end{enumerate}
The Viterbi algorithm is an example of a dynamical programming algorithm, where a computation can be broken down into a series of sub-computations.
The Viterbi path for a time series sequences spanning $t_0$ to $t_{N_t}$ contains the Viterbi path for sub-sequences of that time series. 
It is also very computationally efficient. \jam{[Justify this claim to efficiency? Cite something about it being $O(n^2)$ or whatever it is?]}

\begin{comment}
The route through the grid which maximises the overall probability

Constraints are set on the allowable paths. 
In continuous wave searches, a physical model is used to decide how much the frequency can wander at each timestep, depends on the type of target. 
In this work we allow the frequency to either a) stay in the same bin; b) move up a single frequency bin; or c) move down a single frequency bin. 
We label the probability of transitioning from $f_j$ at $t_{i}$ to $f_{j+1,j,j-1}$ at $t_{i+1}$ as $A_{i,j}$. 
In this work the three transition option have equal probability.

Our objective is now to find the most likely path through the grid given the data and transition constraints. 
This is the route which maximises the overall probability, which we call the `Viterbi path'.
In practice we sum the logarithm of the Fourier amplitude $\log O_{i,j}$ along the path.

The Viterbi algorithm begins at $t_{i=1}$. 
The algorithm selects the highest probability route for each $f_j$. 
\begin{equation}
A_{i,j} O_{i,j+1}
A_{i,j} O_{i,j}
A_{i,j} O_{i,j-1}
\end{equation}
At $t_{i=2}$ it then selects the best route to this point based on the accumulated probability along the paths so far. 
For example, at

(a.) The grid point labelled (a) has $f_{j=3}$ and $t_{i=1}$. 
Looking back at the previous time, there are three options for where the path could have started from at $t_{i=0}$. These are $f_{j=4}$, $f_{j=3}$, or $f_{j=2}$. 
The algorithm selects $f_{j=3}$ as it has the highest value. 
(b.) The grid point labelled (b.) is not the best starting position for any of the cells at $t_{i=1}$, no paths starting from this position will be considered from now on.

The probability of a path is given by the accumulated probability along the path. 
\begin{equation}

\end{equation}

The value of each element in the grid is the logarithm of the normalised Fourier amplitude, which we label $O_{i,j}$.

Our objective is to find the most likely path through the grid which maximises the , the `best path'. 
In practice, we sum the logarithm of the Fourier amplitudes along the path.

We label the time-frequency grid points as $t_i$ and $f_j$ respectively, where $i=1,2,...N_t$ and $j=1,2,...,N_f$ (and $N_t$ and $N_f$ are the total number of time and frequency bins respectively). 
Starting at $t_{i=2}$, the algorithm iterates calculates the best path to each $f_j$ from $t_{i=1}$ to $t_=2$.  through $t_i$, at each stage finding the best path to every $f_j$ from $t_{i=0}$.

Abstractly speaking, the algorithm is given a weighted graph (e.g.\ the spectrogram grid), a sensible sequence of subgraphs (e.g.\ the columns of the spectrum at each time), and restrictions on connectivity (e.g.\ the allowed amount of frequency wander). 
At each iteration, the algorithm will find the best path to each node in the next sub-graph. 
This is repeated to the final sub-graph. 
%n, at each iteration, it will find the best path to each node in the next subgraph, all the way from the first subgraph. 
The Viterbi path is then the overall best path from the first sub-graph to the last. 
The overall best path from the first sub-graph to the last is then selected, which is called the Viterbi path. 
At the end, it selects the overall best path from the first to the last subgraph (here, from the start to end time), this overall best path is called the Viterbi path. 
See Appendix~\ref{app:viterbi} for the full details of the algorithm in this particular implementation.

% moved from above
The method requires a limit for the allowed frequency change over time, usually informed by a physical model. 
In this work the frequency $f$ path can move in one of three ways at each time bin: a) up one frequency bin; b) stay in the same frequency bin; or c) down one frequency bin. 
We set the frequency bin width to be $\Delta f = 0.3\,{\rm Hz}$ which corresponds to a maximum frequency wander of $0.11\,{\rm Hz}{\rm s}^{-1}$.
%In this work, the frequency wander limit is set to $0.11\,{\rm Hz} {\rm s}^{-1}$ which corresponds to one frequency bin per every time bin. 
In some cases, we see this proves poorly adjusted to the signal. 
\end{comment}

@@

The results are shown in Fig.~\ref{fig:viterbi_overlay}, where the heatmap shows the spectrogram of the observed signal. 
The overlaid pink-dot and white-cross markers show the injected signal and recovered Viterbi path respectively maximized over all terminating frequency bins. 
The Viterbi path provides a good match to the injected frequency. 
%To the eye, the Viterbi algorithm easily recovers the injected frequency over time due to the high signal-to-noise ratio (SNR) of this experiment \han{What was the SNR? / can we quantify this?}.
Quantitatively, the Viterbi path is within one frequency bin ($\approx 0.3\,{\rm Hz}$) of the injected signal for $94\%$ of the time. 
We also consider the root mean square (RMS) fractional error along the path. Given $i=0,1,...,N_f$ columns of the injected frequency $I_i$ and recovered (Viterbi) frequency $R_i$, the RMS fractional error is given by:
\begin{equation}
\sqrt{ \frac{1}{N_f+1} \sum_{i=0}^{N_f} \frac{(I_i - R_i)^2}{{I_i}^2} }
\end{equation}
Computing this along the path, we end up with the value \jam{???}, which indicates a good recovery of the injected signal.

>>>>

Continuous gravitational wave signals may wander slowly in frequency over time, due to stochastic internal processes in the superfluid interior of isolated neutron stars~\cite{MelatosDouglassSimula:2015,Jones:2010} or stochastic accretion flows in binary neutron stars~\cite{BildstenTB:1998}. 
%As also mentioned in Section~\ref{sec:introduction}, continuous wave searches often look for signals that wander slowly in frequency~\cite{ScoX1O2Viterbi:2019}. 
The audio analogue of this is a tone that wanders in frequency; a note that changes pitch. 
%Unlike the single tone described in Section~\ref{sec:single_tone}, finding the signal isn’t as simple as searching for a peak in the spectrum for the full timeseries data.%observing run.

@@

@@

\item The Viterbi path is then found by backtracking along the best paths at each time step.
\jam{[Andrew mentioned providing a closed form here, which I'm not sure exists. The code works by storing three arrays: the weights, the score for the best path to a given vertex, and the index of the previous vertex along that path. Backtracking comprises of looping through that third array from the vertex in the last row with the highest score from the second array. I don't think we need to explain it any further than that.]}
In Fig.~\ref{fig:viterbi}, we see that the path ending at (b) started at (c). 
The Viterbi path is highlighted in orange in fig.~\ref{fig:viterbi}.
\end{enumerate}
The Viterbi algorithm is an example of a dynamical programming algorithm, where a computation can be broken down into a series of sub-computations.
The Viterbi path for a time series sequences spanning $t_0$ to $t_{N_t}$ contains the Viterbi path for sub-sequences of that time series. 
The algorithm is computationally efficient, at every time step all but $N_f$ possible paths are eliminated (see also Ref.~\cite{ScoX1ViterbiO1:2017}).

@@

The results are shown in Fig.~\ref{fig:viterbi_overlay}, where the heatmap shows the spectrogram of the observed signal. 
The overlaid pink-dot and white-cross markers show the injected signal and recovered Viterbi path respectively maximized over all terminating frequency bins. 
The Viterbi path provides a good match to the injected frequency. 
%To the eye, the Viterbi algorithm easily recovers the injected frequency over time due to the high signal-to-noise ratio (SNR) of this experiment \han{What was the SNR? / can we quantify this?}.
Quantitatively, the Viterbi path is within one frequency bin ($\approx 0.3\,{\rm Hz}$) of the injected signal for $94\%$ of the time. 
We also consider the root mean square (RMS) fractional error along the path. Given $i=0,1,...,N_f$ columns of the injected frequency $I_i$ and recovered (Viterbi) frequency $R_i$, the RMS fractional error is given by:
\begin{equation}
\sqrt{ \frac{1}{N_f+1} \sum_{i=0}^{N_f} \frac{(I_i - R_i)^2}{{I_i}^2} }
\end{equation}
Computing this along the Viterbi path we find an error of 0.082, which indicates fair but not total recovery of the injected signal.

### 2020-08-16 paper/paper_main.tex
adding app c, general text editing and tidying
tasks 1, 3, and 5
task 4
some general text editing and addressing comments

% Note that AJP uses the same style as Phys. Rev. B (prb).
% The AIP Style Manual\cite{AIPstylemanual} is an indispensable reference on good physics writing, covering everything from planning and organization to standard spellings and abbreviations.
% Most important of all, please familiarize yourself with the AJP Statement of Editorial Policy,\cite{editorsite} which describes the types of manuscripts that AJP publishes and the audience for which AJP authors are expected to write.
% We look forward to receiving your submission to AJP.
\usepackage[utf8]{inputenc}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage{float}
\usepackage{mathtools}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{hyperref}
% \usepackage{siunitx}
\usepackage{textcomp}
\usepackage{subfiles}
\usepackage[bottom]{footmisc}
\usepackage{comment}

@@

@@

\author{William Moran}
\email{wmoran@unimelb.edu.au}
\affiliation{%
Department of Electrical and Electronic Engineering, University of Melbourne, Parkville, Victoria 3010, Australia
}
\affiliation{%
\han{check all affiliations}
}

\author{Changrong Liu}
\email{}
\affiliation{%
Department of Electrical and Electronic Engineering, University of Melbourne, Parkville, Victoria 3010, Australia
}
\affiliation{%
\han{check all affiliations}
}

\author{\han{(to do: ask other authors?)} \jam{[Other authors and affiliations I leave to you to contact and clarify]}}
\email{wmoran@unimelb.edu.au}
\affiliation{%
Adelaide, Adelaide-OzGrav
}

@@

\begin{abstract}
Gravitational wave observatories around the world are searching for continuous waves: persistent signals from spinning neutron stars. 
Searches use sophisticated statistical techniques to look for weak signals in noisy data. 
We show how continuous wave searches can be demonstrated with a table-top model gravitational-wave detector (a Michelson interferometer), where sound plays the role of gravitational waves. 
Using signal processing techniques from continuous wave searches we demonstrate the recovery of tones with constant and wandering frequencies. 
Finally we investigate the performance of the demonstration as an `optical microphone' for recovering music and speech. 
Simple chords and drums are easily recovered, but speech is more challenging. 
%however speech recordings are very noisy. 
%Application of an existing speech enhancement technique provides some noise reduction, however the audio remains unintelligible. 
This demonstration has scope for use in the physics and electrical engineering undergraduate lab. 
It can also be adapted for use as an engagement tool for communicating gravitational-wave and signal-processing topics to non-specialist audiences. 
%Continuous gravitational wave searches use sophisticated statistical techniques to search for weak signals in noisy data. 
%We demonstrate these techniques by observing sound with a table-top Michelson interferometer. 
%In particular, we used the Viterbi algorithm to recover a tone with wandering frequency from a webcam video of the interference pattern.
%We also describe how the inclusion of a photodiode and Raspberry Pi allowed the interferometer to act as an `optical microphone' with sufficient fidelity to play back simple audio recordings. 
%However, more complex recordings, of speech, remain unintelligible when played back, despite preliminary efforts made towards speech enhancement.
%Continuous gravitational wave searches use sophisticated statistical techniques to search for weak signals in noisy interferometric data. 
%We demonstrate these techniques by observing sound with a table-top Michelson interferometer. 
%In particular, we used the Viterbi algorithm to recover a tone with wandering frequency from a webcam video of the interference pattern.
%We also describe how the inclusion of a photodiode and Raspberry Pi allowed the interferometer to act as an `optical microphone' with sufficient fidelity to play back simple audio recordings. 
%However, more complex recordings, of speech, remain unintelligible when played back, despite preliminary efforts made towards speech enhancement.
\end{abstract}

@@

\begin{acknowledgments}
The authors are grateful to Jude Prezens, Alex Tolotchkoc, and Blake Molyneux for their technical advice and generous assistance throughout the project.
The authors are also grateful to Deeksha Beniwal, Sebastian Ng, and Craig Ingram for their advice and work in designing the interferometer.
\jam{[update if turned into authors]} 
This research is supported by the Australian Research Council Centre of Excellence for Gravitational Wave Discovery (OzGrav) (project number CE170100004) and the Institute of Physics International Member Grant.
Travel during the project was supported by the Australian National University PhB Science program.

@@

\newpage
\appendix
\subfile{ifo-appendix.tex}

>>>>

% Note that AJP uses the same style as Phys. Rev. B (prb).
% The AIP Style Manual\cite{AIPstylemanual} is an indispensable reference on good physics writing, covering everything from planning and organization to standard spellings and abbreviations.
% Most important of all, please familiarize yourself with the AJP Statement of Editorial Policy,\cite{editorsite} which describes the types of manuscripts that AJP publishes and the audience for which AJP authors are expected to write.
% We look forward to receiving your submission to AJP.
\usepackage[utf8]{inputenc}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage{float}
\usepackage{mathtools}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{hyperref}
% \usepackage{siunitx}
\usepackage{textcomp}
\usepackage{subfiles}
\usepackage[bottom]{footmisc}
\usepackage{comment}
\usepackage{algpseudocode}
% \usepackage{algorithm}
% \usepackage{algorithmic}

\usepackage{lineno}
\linenumbers

@@

\author{James W. Gardner}
\email{u6069809@anu.edu.au}
\affiliation{%
College of Science, Australian National University, Acton, ACT, 2601, Australia 
}%
\affiliation{%
 School of Physics, University of Melbourne, Parkville, Victoria, 3010, Australia
}

@@

\author{Changrong Liu}
\email{}
\affiliation{%
Department of Electrical and Electronic Engineering, University of Melbourne, Parkville, Victoria 3010, Australia
}

\author{...}
\email{to do}
\affiliation{%
Adelaide, Adelaide-OzGrav
}

@@

\begin{abstract}
Gravitational-wave observatories around the world are searching for continuous waves: persistent signals from spinning neutron stars. 
Searches use sophisticated statistical techniques to look for weak signals in noisy data. 
We show how continuous-wave searches can be demonstrated with a table-top model gravitational-wave detector (a Michelson interferometer), where sound plays the role of gravitational waves. 
Using signal processing techniques from continuous-wave searches we demonstrate the recovery of tones with constant and wandering frequencies. 
Finally we investigate the performance of the demonstration as an `optical microphone' for recovering music and speech. 
Simple chords and drums are easily recovered, but speech is more challenging. 
%however speech recordings are very noisy. 
%Application of an existing speech enhancement technique provides some noise reduction, however the audio remains unintelligible. 
This demonstration has scope for use in the physics and electrical engineering undergraduate lab. 
It can also be adapted for use as an engagement tool for communicating gravitational-wave and signal-processing topics to non-specialist audiences. 
%Continuous gravitational wave searches use sophisticated statistical techniques to search for weak signals in noisy data. 
%We demonstrate these techniques by observing sound with a table-top Michelson interferometer. 
%In particular, we used the Viterbi algorithm to recover a tone with wandering frequency from a webcam video of the interference pattern.
%We also describe how the inclusion of a photodiode and Raspberry Pi allowed the interferometer to act as an `optical microphone' with sufficient fidelity to play back simple audio recordings. 
%However, more complex recordings, of speech, remain unintelligible when played back, despite preliminary efforts made towards speech enhancement.
%Continuous gravitational wave searches use sophisticated statistical techniques to search for weak signals in noisy interferometric data. 
%We demonstrate these techniques by observing sound with a table-top Michelson interferometer. 
%In particular, we used the Viterbi algorithm to recover a tone with wandering frequency from a webcam video of the interference pattern.
%We also describe how the inclusion of a photodiode and Raspberry Pi allowed the interferometer to act as an `optical microphone' with sufficient fidelity to play back simple audio recordings. 
%However, more complex recordings, of speech, remain unintelligible when played back, despite preliminary efforts made towards speech enhancement.
\end{abstract}

@@

\begin{acknowledgments}
The authors are grateful to Jude Prezens, Alex Tolotchkoc, and Blake Molyneux for their technical advice and generous assistance throughout the project. 
This research is supported by the Australian Research Council Centre of Excellence for Gravitational Wave Discovery (OzGrav) (project number CE170100004). 
This work also received financial support from the Institute of Physics International Member Grant and travel support from the Australian National University PhB Science program.

@@

\appendix
\subfile{ifo-appendix.tex}

\nocite{*}
\bibliographystyle{myunsrt}
\bibliography{ifoDemoBib}
